Using reference models in variable selection \vspace{.1in}

### 2020-04-27 paper/ref_approach.tex
minor edits

\begin{document}
\maketitle
\begin{abstract}
  Variable selection, or more generally, model reduction is an important aspect of the statistical workflow aiming to provide insights from data. In this paper, we discuss and demonstrate the benefits of using a reference model in variable selection. A reference model acts as a noise-filter on the target variable by modeling its data generating mechanism. As a result, using the reference model predictions in the model selection procedure reduces the variability and improves stability leading to improved model selection performance. Assuming that a Bayesian reference model describes the true distribution of future data well, the theoretically preferred usage of the reference model is to project its predictive distribution to a reduced model leading to projection predictive variable selection approach. Alternatively, reference models may also be used in an ad-hoc manner in combination with common variable selection methods.  In several numerical experiments, we investigate the performance of the projective prediction approach as well as alternative variable selection methods with and without reference models. Our results indicate that the use of reference models generally translates into better and more stable variable selection. Additionally, we demonstrate that the projection predictive approach shows superior performance as compared to alternative variable selection methods independently of whether or not they use reference models.
% These benefits are also illustrated on a real-world example of predicting the amount of body fat for which we compare a reference model based selection method to a purely data driven approach.
\end{abstract}

@@

To motivate the further discussion and experiments, we start by a simple
variable selection example using body fat data by \citet{johnson1996fitting}.
We compare the projective prediction approach (\emph{projpred},
\citet{paper:projpred}) which uses a reference model, and classic
stepwise backward regression (\emph{steplm}).
% \footnote{The analysis is based on \texttt{R}-notebook by Vehtari available at
%   \url{https://avehtari.github.io/modelselection/bodyfat.html}.}.
The experiments are implemented in \texttt{R} \citep{Rcore2018}.

@@

\cite{paper:bodyfat} report results using steplm with a significance
level of 0.157 with AIC selection \citep{akaike1974new}, fixing
abdomen and height to be always included in the model. For better
comparison we do not fix any of the variables.  The steplm approach is
carried out combining the \texttt{step} and \texttt{lm} functions in R.

For the selection via projpred, the Bayesian reference model includes
all the variables using a regularised horseshoe prior~\citep{paper:rhs} on the
variable coefficients. Submodels are explored using
forward search (the results are not sensitive to whether forward or
backward search is used), and the predictive utility is the expected
log-predictive density (elpd) estimated using approximate leave-one-out cross-validation via Pareto-smoothed importance-sampling
\citep[PSIS-LOO-CV; ][]{paper:psis_loo}.  We select the smallest  submodel with an elpd score similar to the reference model
when taking into account the uncertainty in estimating the predictive
model performance. See Appendix A for brief review of projection
predictive approach, and papers by \citet{paper:model_selection} and
\citet{paper:projpred} for more details. The complete projpred approach is implemented in the \texttt{projpred} R package \citep{Rprojpred}.

@@

\begin{table}[tp]
\footnotesize
\centering
\begin{tabular}{l|lr|lr}
%  \hline
M & projpred & Freq \% & steplm & Freq \%  \\ 
  \hline
1 & abdom., weight & 39 & abdom., age, forearm, height, hip, neck, thigh, wrist & 4 \\
2 & abdom., wrist & 10 & abdom., age, chest, forearm, height, neck, thigh, wrist & 4 \\
3 & abdom., height & 10 & abdom., forearm, height, neck, wrist & 2 \\
4 & abdom., height, wrist & 9 & abdom., forearm, neck, weight, wrist & 2 \\
5 & abdom., weight, wrist & 8 & abdom., age, height, hip, thigh, wrist & 2 \\
6 & abdom., chest, height, wrist & 2 & abdom., age, height, hip, neck, thigh, wrist & 2 \\
7 & abdom., biceps, weight, wrist & 2 & abdom., age, ankle, forearm, height, hip, neck, thigh, wrist & 2 \\
8 & abdom., height, weight, wrist & 2 & abdom., age, biceps, chest, height, neck, wrist & 2 \\
9 & abdom., age, wrist & 2 & abdom., age, biceps, chest, forearm, height, neck, thigh, wrist & 2 \\
10 & abdom., age, height, neck, thigh, wrist & 2 & abdom., age, ankle, biceps, weight, wrist & 2 \\
%   \hline
\end{tabular}
\caption{Body fat example: Bootstrap model selection frequencies from 100 bootstrap samples.  The projpred approach has less variability on which variable combinations are selected.}
\label{tab:model_frequencies}
\end{table}

@@

We repeat the experiment with a modified data set by adding 84
unrelated noisy variables so that there are 100 variables in total. 
The last two rows of Table
\ref{tab:model_performances} show the cross-validated RMSE, the size
of the selected model and the number of selected noisy variables using
projpred or steplm. The results show that projpred has similar
predictive performance and the same number of selected variables as
with the original data, whereas the stepwise regression has worse
predictive performance and the number of selected variables is much
higher and include a large number of irrelevant variables.

\begin{table}[tp]
\footnotesize
\centering
\begin{tabular}{ll|rrrr}
%  \hline
Data  & Method & RMSE Full & RMSE Sel & \# Sel (10-CV avg, sd) & \# Sel noisy (10-CV avg, sd) \\ 
  \hline
body fat & projpred & 4.4 & 4.5 &  2\, (2.3, 0.5) &   \\
& steplm & 4.4 & 4.5 & 7\, (6.0, 0.9) &  \\
\hline
+ noisy variables & projpred & 4.5 & 4.5 &  2\, (2.0, 0.0)&  0\,\; (0,\, 0)  \\
& steplm & 5.7 & 5.1 & 23\, (26,\; 4.5) &  15\, (19, 4) \\
%   \hline
\end{tabular}
\caption{Body fat example: Predictive performances with original data (first two
  rows) and with extra noisy variables (last two rows) estimated with
  10-fold cross-validation. Abbreviations: RMSE = root mean squared error;
  Full = full model; Sel = selected submodel; 
  \# Sel = total number of selected variables;
  \# Sel noisy = number of selected noisy variables. 
  10-CV avg = average over the 10 folds in cross-validation. 10-CV sd = standard deviation
  over the 10 folds in cross-validation.}
\label{tab:model_performances}
\end{table}

@@

\hypertarget{reference-model-approach}{%
\subsection{Why the reference model helps}\label{reference-model-approach}}

@@

Using the data generating mechanism \eqref{eq:simulated_data}, we
simulate data sets of different sizes with a relatively large number
of variables $p=70$, with $k=20$ of them being predictive.  We compare
the minimal subset variable selection performance of the projection
predictive approach (which uses a reference model and it is referred to as projpred), 
a Bayesian stepwise
forward selection with and without a reference model, and maximum
likelihood stepwise forward selection with and without a reference
model (steplm). The following is a summary of the implementation of the compared methods:
\begin{itemize}
\item projpred: The projective prediction approach is used. 
  The reference model is a Bayesian linear regression
  model using the first five supervised principal components \citep{piironen2018} as predictors
  and the full posterior predictive distribution as the basis of the projection.   
  The search heuristic  is forward search and the predictive performance 
  is estimated via 10-fold cross-validation. The selection continues
  until the predictive performance is close to the predictive
  performance of the reference model. See more details in Appendix A.

\item Bayesian stepwise selection (without a reference model): At each step, the fitted model
 is a Bayesian linear regression using the regularised horseshoe
  prior and the variable excluded is the one with the highest Bayesian
  $p$-value defined as $\text{min}\{P(\theta\leq0 \; | \; D),P(\theta>0 \; | \; D)\}$,
  where $D$ stays for the observed data.
The selection continues if the reduced model has an elpd score higher 
(i.e., better) than the current model.

\item Bayesian stepwise selection (with a reference model): The reference model is the same as for projpred, but only point predictions (posterior predictive means) $\hat{y}$ are used to replace the target variable $y$. The same Bayesian
  $p$-value selection strategy as in the data based Bayesian stepwise selection is used.

%Same as the data based Bayesian stepwise selection, but instead of using the observed data values to fit the models at each step, it uses the target values predicted by the reference model (which is the same as in the projpred approach).
\item steplm (without a reference model): Stepwise selection using AIC as in the body fat example.

\item steplm (with a reference model): The reference model is the same as for
 projpred, but only point predictions (posterior predictive means) $\hat{y}$ 
 are used to replace the target variable $y$. The same AIC selection strategy 
 as in the data based steplm is used.

@@

When the variable selection is repeated with different simulated data
sets, there is some variability in the selected variables. We
measure the stability of variable selection by computing the entropy
of the observed distribution of the included variables over different models. The smallest entropy
would be obtained if the approach always selected the same set of
variables, and the largest entropy would be observed if the approach
would always selected different sets of variables. Therefore, lower entropy 
corresponds to a more stable selection. Highly correlated predictive variables may happen to be
selected alternately, thus making stability estimation of the
selection a non-trivial task. Entropy can not distinguish the interchangeability 
due to correlation from instability. Thus, such a measure should be 
considered as a relative, and not as an absolute, measure of stability.
Figure \ref{fig:entropy} shows the entropy scores for the different compared
methods. The use of a reference model
improves the stability of steplm in variable selection slightly, while it makes
little difference for the Bayesian linear regression. The
projpred approach turns out to be far more stable than all
other methods. This is likely due to projpred being based on better
decision theoretical formulation which 1) takes into account the full
predictive distribution and not just point estimate and 2) projects
the reference model posterior to the submodel instead of using a
simple refit of submodels.
% TODO: Bigger and more distinctive symbols in figures
% label order: steplm, bayes.step, projpred
\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/rmse_vs_fdr_parallel.pdf}
  \vspace{-0.3\baselineskip}
  \caption{Simulation study 1: Root mean square error (RMSE) against false discovery rate in the minimal subset variable selection with one standard deviation error bars. The projpred approach has the smallest RMSE and false discovery ratio.}
  \label{fig:rmse_vs_fdr}
\end{figure}
% TODO: Bigger and more distinctive symbols in figures
% TODO: order steplm, bayes.step, projpred
\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/entropy_parallel.pdf}
  \vspace{-0.3\baselineskip}
  \caption{Simulation study 1: Entropy score in the minimal subset variable selection.  The projpred approach has much smaller entropy score than the other approaches.}
  \label{fig:entropy}
\end{figure}

\hypertarget{bodyfat-2}{%
\subsection{Body fat example: Part 2}\label{bodyfat-2}}

Here we repeat the selection of Section \ref{bodyfat-1} via
stepwise backward regression. In this case, the overall number of
variables (original plus noisy) is 100, as it was in the last part of
Section \ref{bodyfat-1}. We compare results with and without using
a simple reference model approach outlined in \eqref{eq:ref_mod} with
steplm. Figure \ref{fig:bodyfat_step_refvsdata} shows the number of
irrelevant variables included in the final model and the out-of-sample
root mean square error (RMSE). Results are based on 100 bootstrap
samples on the whole dataset, and the predictive performance is tested
on the observations excluded at each bootstrap sample. We observe that
the reference model reduces the number of irrelevant variables included
in the final model. This leads to less overfitting and thus to
improved out-of-sample predictive performance in terms of RMSE. The
reference model approach applied to the stepwise backward regression
achieves outstanding improvements considering its simplicity, yet it
does not reach the goodness of the much more sophisticated projective
prediction approach (see results of Section \ref{bodyfat-1}).
\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/bodyfat_step_refvsdata.pdf}
  \vspace{-0.7\baselineskip}
  \caption{Body fat example: Stepwise backward selection with and without using a reference model. The x-axis denotes the number of selected irrelevant variables on the left and the out-of-sample RMSE on the right-hand side based on 100 bootstrap samples. The reference approach reduces the number of noisy variables selected and the out-of-sample RMSE.}
  \label{fig:bodyfat_step_refvsdata}
\end{figure}
\begin{table}[tp]
\footnotesize
\centering
\begin{tabular}{l|rr}
%  \hline
Method & RMSE & Selected noisy \\ 
  \hline
Stepwise selection & 6.9 (0.7) & 48 (7)    \\
Reference model + Stepwise selection & 4.7 (0.5) & 25 (7)   \\
\end{tabular}
\caption{Body fat example: Means and standard deviations (between brackets) of the results shown in Figure \ref{fig:bodyfat_step_refvsdata}.}
\label{tab:step_refvsdata}
\end{table}

% \hypertarget{comparison}{%
%   \section{A comparison in the normal means problem framework}\label{comparison}}
\section{Complete variable selection}\label{comparison-complete-subset}
\label{complete-selection}

@@

%Table \ref{tab:comparison} summarises the methods used in the numerical experiments.

@@

\begin{center}
\begin{algorithm}[H]
\SetAlgoLined
\KwResult{$R:=\{\text{selected variables}\}$}
 $F:=\{\text{set of variables}\}$ \;
 $R=\{\emptyset\}$ \;
 Fit reference model\;
 \While{$F\neq\{\emptyset\}$}{
   projection.HeuristicSearch() \;
   projection.elpdEstimate() \;
%:
   $S=\text{min}\{\text{sub}: P(\text{elpd}_{\text{sub}}-\text{elpd}_{\text{base}}>0)\geq\alpha)$\} \;

@@

% Many Bayesian approaches for complete variable selection first fit a
% Bayesian regression model using some sparsifying prior on the
% regression coefficients, for example, as discussed in
% \cite{paper:dirichlet_laplace}, \cite{bhadra2017horseshoe+} and
% \cite{johnstone2004needles}.  In our experiments, we include the
% Dirichlet-Laplace \cite[DL;][]{paper:dirichlet_laplace} prior and the
% regularised horseshoe (RHS) prior \citep{paper:rhs}, which in our experience
% scales better to higher number of variables as compared to other
% horseshoe-like priors.
% \note{Aki: How this paragraph connects to the next paragraph?}

% If these priors have continuous support, as is the case for the DL and
% horseshoe-like priors \citep{paper:hs,bhadra2017horseshoe+,paper:rhs},
% there is no canonical way to complete the selection since the
% posterior probability of an exactly zero signal is always
% zero. Whereas \cite{paper:dirichlet_laplace} used k-means clustering
% over the posterior samples to perform the actual selection,
% \cite{bhadra2017horseshoe+} just focused their analysis on the
% obtained shrinkage and the correct estimation of the sparse signals
% without any actual selection.

% For the Bayesian regression
% models with these priors, we compare the posterior shrinkage and the
% correct estimation of the sparse signals with and without using a
% reference model as a noise filter. For this purpose, we consider the
% average SSE (sum of square errors) and the average SESE (sum of
% expected square errors) as quantities measuring estimation
% accuracy. They are defined as \
% \begin{align}
% \text{SSE}&=\frac{1}{n-3}\sum_{j=1}^{p}(\hat{\theta}_{j} - \theta^{0}_{j})^{2} \label{eq:SSE} \\
% \text{SESE}&=\frac{1}{n-3}\sum_{j=1}^{p}\int_{\mathbb{R}}(\theta_{j}-\theta^{0}_{j})^{2}p(\theta_{j}|z_{1},\ldots,z_{p})d\theta_{j} \\
% &=\frac{1}{n-3}\sum_{j=1}^{p}\mathbb{E}_{\theta_{j}}[(\theta_{j}-\theta^{0}_{j})^{2}|z_{1},\ldots,z_{p}], \label{eq:SESE}
% \end{align}
% where $\theta_{j}^{0}=\sqrt{n-3}\,T_{F}(\rho_{j})$ is the true value,
% $\hat{\theta}_{j}$ is the posterior median of each parameter
% $\theta_{j}$. We divide the errors by $(n-3)$ to have quantities
% independent of the number of statistical units.  The SESE takes the
% posterior uncertainty in account and thus also the shrinkage of the
% posterior distributions, whereas the SSE only measures the bias in the
% point estimates.  Any (sensible) point estimate could be used to
% compute the SSE but here we follow \cite{paper:dirichlet_laplace} and
% use the posterior median. The results that we show later on in Section
% \ref{shrinkage-signal} are not sensitive to the choice of this point
% estimate.

% For shrinkage priors, one central aspect is the amount of shrinkage
% they actually impose on the relevant and irrelevant variables. Ideally,
% the irrelevant variables should be completely shrunk to zero while the
% relevant variables should not be shrunk at all. The amount of shrinkage
% can be quantified by the shrinkage factor $\kappa_{j}$ for each
% parameter $\theta_{j}$ ranging from $0$ (no shrinkage) to $1$
% (complete shrinkage). For model \eqref{eq:normal_means_problem2} with
% a hierarchical shrinkage prior of the type
% \begin{align}
% Z_{j}|&\theta_{j}\overset{ind}{\sim}N(\theta_{j},1) \qquad j=1,\ldots,p \\
% &\theta_{j}|\tau,\nu_{j}\overset{ind}{\sim}N(0,\nu_{j}^{2}\tau^{2})
% \end{align}
% the shrinkage factor $\kappa_{j}$ is defined as
% \begin{equation}
% \kappa_{j}=\frac{1}{1+\tau^{2}\nu_{j}^{2}}.
% \end{equation}
% \
% For RHS and DL priors we have
% \begin{equation}
% \kappa_{j}^{\text{RHS}}=\frac{1}{1+\tau^{2}\lambda_{j}^{2}}, \qquad \kappa_{j}^{\text{DL}}=\frac{1}{1+\psi\phi_{j}^{2}\tau^{2}}.
% \end{equation}
% In the left hand-side expression, $\tau$ and $\lambda_{j}$ represent
% the global and local shrinkage parameters of the regularised horseshoe
% prior \cite[for further details see][]{paper:rhs}, whereas on the
% right hand-side, $\psi\phi_{j}^{2}\tau^{2}$ is the variance term of
% the Dirichlet-Laplace prior of parameter $\theta_{j}$ following the
% notation of \cite{paper:dirichlet_laplace}. We use the posterior mean
% of $\kappa_{j}$ as a point estimate.

%Within the set of complete selection procedures, we consider three
%Bayesian methods.  \cite{johnstone2004needles} use a spike-and-slab
%type prior given by a mixture of a delta distribution in zero and a
%heavy-tailed distribution \cite[see
%also][]{paper:spike_slab_mitchell}. When the latter is a Laplace
%distribution, they show an interesting thresholding property of the
%median estimator and, based on this observation, present a complete
%selection procedure using their prior. We refer to this method as
%\emph{EB.med}. Further, we use a method to control for the local false
%discovery rate (\emph{loc.fdr}) as described in \cite{paper:efron,
%  efron2012large}. Finally, we device a simple selection procedure
%based on posterior credible intervals using the regularised horseshoe
%prior in which we select variables whose 90\% credible intervals
%excludes zero (\emph{ci.90}) \note{Reference for this?}.

@@

%\note{Aki: ``Type: Complete selection'' doesn't make sense in the table. Maybe we need to discuss when to use the term complete selection and what is the alternative title for this section.}

% \begin{table}[tp]
% \scriptsize
% \centering
% \begin{tabular}{l|l|l|l}
% Method & Abbreviation & Type & Quantities of interest \\ 
%  \hline
% Regularised Horseshoe & RHS & Shrinkage prior & SSE, SESE, $\kappa$ \\
% Dirichlet-Laplace & DL & Shrinkage prior & SSE, SESE, $\kappa$ \\
% Local false discovery rate & loc.fdr & Complete selection & fdr, sensitivity, stability \\
% Median estimator & EB.med & Complete selection & fdr, sensitivity, stability \\
% Credible intervals & ci.90 & Complete selection & fdr, sensitivity, stability \\
% \end{tabular}
% \caption{Summary of the shrinkage and selection methods used in the numerical experiments.}
% \label{tab:comparison}
% \end{table}

@@

% The normal means problem formulation is obtained as explained in
% Section \ref{comparison} by applying the Fisher-z-transformation to
% the empirical correlations. We label the reference approach as
% ``ref'', which consists of using the reference model on top of the
% computation of the z-values, and the standard approach based on the
% observed data simply as ``data''. In Section \ref{shrinkage-signal},
% we analyse the different shrinkage and signal estimates in the context
% of the full Bayes estimation methods by comparing the ref and data
% approach. In Section \ref{complete-selection}, we present results for
% the complete selection procedures.

% \hypertarget{shrinkage-signal}{%
% \subsubsection{Shrinkage and signal analysis}\label{shrinkage-signal}}
% In the framework of fully Bayesian estimation methods, we analyse the
% effect of the reference model using two continuous hierarchical
% shrinkage priors: the Dirichlet-Laplace
% \citep{paper:dirichlet_laplace} and the regularised horseshoe prior
% \citep{paper:rhs}.
% %
% For the Dirichlet hyperparameter of the Dirichlet-Laplace prior, we
% use an uniform prior on the interval $(1/p,1/2)$. For the regularised
% horseshoe prior, we follow the recommendations provided in
% \cite{paper:rhs} by setting the prior guess of the number of relevant
% variables to one in order to achieve the highest possible
% shrinkage. The results are not sensitive with respect to the specific
% choice.  \ Since, as explained at the start of Section
% \ref{comparison}, continuous priors usually do not provide an
% automatic selection procedure, we study the amount of shrinkage and
% the bias of the posterior medians looking at the SESE and the SSE (see
% definitions \eqref{eq:SSE} and \eqref{eq:SESE}), together with the
% posterior mean of the shrinkage factors.

% \begin{figure}[tp]
%   \centering
%   \includegraphics[width=0.98\textwidth]{graphics/post_int.pdf}
%   \caption{Posterior intervals for the parameters of the normal means
%     problem ($n=50$, $\rho=0.3$) using the regularised horseshoe
%     prior. Posterior means and one standard deviation intervals are
%     depicted. Relevant and non-relevant variables are displayed in red
%     and in blue respectively. Horizontal dashed lines corresponds to
%     the true value of the mean for the relevant variables.\\}
%   \label{fig:posterior_intervals}
% \end{figure}

% Figure \ref{fig:posterior_intervals} shows an example of posterior
% distributions for the model \eqref{eq:normal_means_problem2} using the
% regularised horseshoe prior with the data approach on the left and the
% ref approach on the right-hand side. It is evident how the reference
% model helps in tearing apart the signal from the noise, resulting in
% less biased estimates and clearer separation between relevant and
% irrelevant variables. This is quantified in Figure \ref{fig:SESE_SSE}
% where average SESE and SSE are depicted based on 100 simulations. We
% observe that the reference approach (in orange) has a clearly lower
% estimate for both the SESE and the SSE. As $n$ and $\rho$ increase,
% the amount of error diminishes, as expected, due to a larger amount of
% information of the relevance of each variable provided by the
% data. The two priors (RHS and DL) seem to achieve similar results
% under all conditions.

% \begin{figure}[tp]
%   \centering
%   \includegraphics[width=0.98\textwidth]{graphics/SESE_SSE.pdf}
%   \caption{Average SESE and SSE with one standard deviation intervals based on 100 simulations.\\}
%   \label{fig:SESE_SSE}
% \end{figure}

>>>>

\begin{document}
\maketitle
\begin{abstract}
  Variable selection, or more generally, model reduction is an important aspect of the statistical workflow aiming to provide insights from data. In this paper, we discuss and demonstrate the benefits of using a reference model in variable selection. A reference model acts as a noise-filter on the target variable by modeling its data generating mechanism. As a result, using the reference model predictions in the model selection procedure reduces the variability and improves stability leading to improved model selection performance. Assuming that a Bayesian reference model describes the true distribution of future data well, the theoretically preferred usage of the reference model is to project its predictive distribution to a reduced model leading to projection predictive variable selection approach. Alternatively, reference models may also be used in an ad-hoc manner in combination with common variable selection methods.  In several numerical experiments, we investigate the performance of the projective prediction approach as well as alternative variable selection methods with and without reference models. Our results indicate that the use of reference models generally translates into better and more stable variable selection. Additionally, we demonstrate that the projection predictive approach shows superior performance as compared to alternative variable selection methods independently of whether or not they use reference models.
\end{abstract}

@@

To motivate the further discussion and experiments, we start by a simple
variable selection example using body fat data by \citet{johnson1996fitting}.
We compare the projective prediction approach (\emph{projpred},
\citet{paper:projpred}) which uses a reference model, and classic
stepwise backward regression (\emph{steplm}).
The experiments are implemented in \texttt{R} \citep{Rcore2018}.

@@

\cite{paper:bodyfat} report results using steplm with a significance
level of 0.157 with AIC selection \citep{akaike1974new}, fixing
abdomen and height to be always included in the model. For better
comparison we do not fix any of the variables.  The steplm approach is
carried out combining the \texttt{step} and \texttt{lm} functions in \texttt{R}.

For the selection via projpred, the Bayesian reference model includes
all the variables using a regularised horseshoe prior~\citep{paper:rhs} on the
variable coefficients. Submodels are explored using
forward search (the results are not sensitive to whether forward or
backward search is used), and the predictive utility is the expected
log-predictive density (elpd) estimated using approximate leave-one-out cross-validation via Pareto-smoothed importance-sampling
\citep[PSIS-LOO-CV; ][]{paper:psis_loo}.  We select the smallest  submodel with an elpd score similar to the reference model
when taking into account the uncertainty in estimating the predictive
model performance. See Appendix A for brief review of projection
predictive approach, and papers by \citet{paper:model_selection} and
\citet{paper:projpred} for more details. The complete projpred approach is implemented in the \texttt{projpred} \texttt{R} package \citep{Rprojpred}.

@@

\begin{table}[tp]
\footnotesize
\centering
\begin{tabular}{l|lr|lr}
M & projpred & Freq \% & steplm & Freq \%  \\ 
  \hline
1 & abdom., weight & 39 & abdom., age, forearm, height, hip, neck, thigh, wrist & 4 \\
2 & abdom., wrist & 10 & abdom., age, chest, forearm, height, neck, thigh, wrist & 4 \\
3 & abdom., height & 10 & abdom., forearm, height, neck, wrist & 2 \\
4 & abdom., height, wrist & 9 & abdom., forearm, neck, weight, wrist & 2 \\
5 & abdom., weight, wrist & 8 & abdom., age, height, hip, thigh, wrist & 2 \\
6 & abdom., chest, height, wrist & 2 & abdom., age, height, hip, neck, thigh, wrist & 2 \\
7 & abdom., biceps, weight, wrist & 2 & abdom., age, ankle, forearm, height, hip, neck, thigh, wrist & 2 \\
8 & abdom., height, weight, wrist & 2 & abdom., age, biceps, chest, height, neck, wrist & 2 \\
9 & abdom., age, wrist & 2 & abdom., age, biceps, chest, forearm, height, neck, thigh, wrist & 2 \\
10 & abdom., age, height, neck, thigh, wrist & 2 & abdom., age, ankle, biceps, weight, wrist & 2 \\
\end{tabular}
\caption{Body fat example: Bootstrap model selection frequencies from 100 bootstrap samples.  The projpred approach has less variability on which variable combinations are selected.}
\label{tab:model_frequencies}
\end{table}

@@

We repeat the experiment with a modified data set by adding 84
unrelated noisy variables, resulting in 100 variables in total. 
The last two rows of Table
\ref{tab:model_performances} show the cross-validated RMSE, the size
of the selected model and the number of selected noisy variables using
projpred or steplm. The results show that projpred has similar
predictive performance and the same number of selected variables as
with the original data, whereas the stepwise regression has worse
predictive performance and the number of selected variables is much
higher and include a large number of irrelevant variables.

\begin{table}[tp]
\footnotesize
\centering
\begin{tabular}{ll|rrrr}
Data  & Method & RMSE Full & RMSE Sel & \# Sel (10-CV avg, sd) & \# Sel noisy (10-CV avg, sd) \\ 
  \hline
body fat & projpred & 4.4 & 4.5 &  2\, (2.3, 0.5) &   \\
& steplm & 4.4 & 4.5 & 7\, (6.0, 0.9) &  \\
\hline
+ noisy variables & projpred & 4.5 & 4.5 &  2\, (2.0, 0.0)&  0\,\; (0,\, 0)  \\
& steplm & 5.7 & 5.1 & 23\, (26,\; 4.5) &  15\, (19, 4) \\
\end{tabular}
\caption{Body fat example: Predictive performances with original data (first two
  rows) and with extra noisy variables (last two rows) estimated with
  10-fold cross-validation. Abbreviations: RMSE = root mean squared error;
  Full = full model; Sel = selected submodel; 
  \# Sel = total number of selected variables;
  \# Sel noisy = number of selected noisy variables. 
  10-CV avg = average over the 10 folds in cross-validation. 10-CV sd = standard deviation
  over the 10 folds in cross-validation.}
\label{tab:model_performances}
\end{table}

@@

\hypertarget{reference-model-approach}{
\subsection{Why the reference model helps}\label{reference-model-approach}}

@@

Using the data generating mechanism \eqref{eq:simulated_data}, we
simulate data sets of different sizes with a relatively large number
of variables $p=70$, with $k=20$ of them being predictive.  We compare
the minimal subset variable selection performance of the projection
predictive approach (which uses a reference model and it is referred to as projpred), 
a Bayesian stepwise
forward selection with and without a reference model, and maximum
likelihood stepwise forward selection with and without a reference
model (steplm). The following is a summary of the implementation of the compared methods:
\begin{itemize}
\item \textbf{projpred}: the projective prediction approach is used. 
  The reference model is a Bayesian linear regression
  model using the first five supervised principal components \citep{piironen2018} as predictors
  and the full posterior predictive distribution as the basis of the projection.   
  The search heuristic  is forward search and the predictive performance 
  is estimated via 10-fold cross-validation. The selection continues
  until the predictive performance is close to the predictive
  performance of the reference model. See more details in Appendix A.

\item \textbf{Bayesian stepwise selection (without a reference model)}: at each step, the fitted model
 is a Bayesian linear regression using the regularised horseshoe
  prior and the variable excluded is the one with the highest Bayesian
  $p$-value defined as $\text{min}\{P(\theta\leq0 \; | \; D),P(\theta>0 \; | \; D)\}$,
  where $D$ stays for the observed data.
The selection continues if the reduced model has an elpd score higher 
(i.e., better) than the current model.

\item \textbf{Bayesian stepwise selection (with a reference model)}: the reference model is the same as for projpred, but only point predictions (posterior predictive means) $\hat{y}$ are used to replace the target variable $y$. The same Bayesian
  $p$-value selection strategy as in the data based Bayesian stepwise selection is used.

\item \textbf{steplm (without a reference model)}: stepwise selection using AIC as in the body fat example.

\item \textbf{steplm (with a reference model)}: the reference model is the same as for
 projpred, but only point predictions (posterior predictive means) $\hat{y}$ 
 are used to replace the target variable $y$. The same AIC selection strategy 
 as in the data based steplm is used.

@@

When the variable selection is repeated with different simulated data
sets, there is some variability in the selected variables. We
measure the stability of variable selection by computing the entropy
of the observed distribution of the included variables over different models. The smallest entropy
would be obtained if the approach always selected the same set of
variables, and the largest entropy would be observed if the approach
would always selected different sets of variables. Therefore, lower entropy 
corresponds to a more stable selection. Highly correlated predictive variables may happen to be
selected alternately, thus making stability estimation of the
selection a non-trivial task. Entropy can not distinguish the interchangeability 
due to correlation from instability. Thus, such a measure should be 
considered as a relative, and not as an absolute, measure of stability.
Figure \ref{fig:entropy} shows the entropy scores for the different compared
methods. The use of a reference model
improves the stability of steplm in variable selection slightly, while it makes
little difference for the Bayesian linear regression. The
projpred approach turns out to be far more stable than all
other methods. This is likely due to projpred being based on better
decision theoretical formulation which 1) takes into account the full
predictive distribution and not just point estimate and 2) projects
the reference model posterior to the submodel instead of using a
simple refit of submodels.

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/rmse_vs_fdr_parallel.pdf}
  \vspace{-0.3\baselineskip}
  \caption{Simulation study 1: Root mean square error (RMSE) against false discovery rate in the minimal subset variable selection with one standard deviation error bars. The projpred approach has the smallest RMSE and false discovery ratio.}
  \label{fig:rmse_vs_fdr}
\end{figure}

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/entropy_parallel.pdf}
  \vspace{-0.3\baselineskip}
  \caption{Simulation study 1: Entropy score in the minimal subset variable selection.  The projpred approach has much smaller entropy score than the other approaches.}
  \label{fig:entropy}
\end{figure}

\hypertarget{bodyfat-2}{
\subsection{Body fat example: Part 2}\label{bodyfat-2}}

Here we repeat the selection of Section \ref{bodyfat-1} via
stepwise backward regression. In this case, the overall number of
variables (original plus noisy) is 100, as it was in the last part of
Section \ref{bodyfat-1}. We compare results with and without using
a simple reference model approach outlined in \eqref{eq:ref_mod} with
steplm. Figure \ref{fig:bodyfat_step_refvsdata} shows the number of
irrelevant variables included in the final model and the out-of-sample
root mean square error (RMSE). Results are based on 100 bootstrap
samples on the whole dataset, and the predictive performance is tested
on the observations excluded at each bootstrap sample. We observe that
the reference model reduces the number of irrelevant variables included
in the final model. This leads to less overfitting and thus to
improved out-of-sample predictive performance in terms of RMSE. The
reference model approach applied to the stepwise backward regression
achieves outstanding improvements considering its simplicity, yet it
does not reach the goodness of the much more sophisticated projective
prediction approach (see results of Section \ref{bodyfat-1}).
\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/bodyfat_step_refvsdata.pdf}
  \vspace{-0.7\baselineskip}
  \caption{Body fat example: Stepwise backward selection with and without using a reference model. The x-axis denotes the number of selected irrelevant variables on the left and the out-of-sample RMSE on the right-hand side based on 100 bootstrap samples. The reference approach reduces the number of noisy variables selected and the out-of-sample RMSE.}
  \label{fig:bodyfat_step_refvsdata}
\end{figure}
\begin{table}[tp]
\footnotesize
\centering
\begin{tabular}{l|rr}
Method & RMSE & Selected noisy \\ 
  \hline
Stepwise selection & 6.9 (0.7) & 48 (7)    \\
Reference model + Stepwise selection & 4.7 (0.5) & 25 (7)   \\
\end{tabular}
\caption{Body fat example: Means and standard deviations (between brackets) of the results shown in Figure \ref{fig:bodyfat_step_refvsdata}.}
\label{tab:step_refvsdata}
\end{table}

\section{Complete variable selection}\label{comparison-complete-subset}
\label{complete-selection}

@@

@@

\begin{center}
\begin{algorithm}[H]
\SetAlgoLined
\KwResult{$R:=\{\text{selected variables}\}$}
 $F:=\{\text{set of variables}\}$ \;
 $R=\{\emptyset\}$ \;
 Fit reference model\;
 \While{$F\neq\{\emptyset\}$}{
   projection.HeuristicSearch() \;
   projection.elpdEstimate() \;
   $S=\text{min}\{\text{sub}: P(\text{elpd}_{\text{sub}}-\text{elpd}_{\text{base}}>0)\geq\alpha)$\} \;

@@

@@

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/sensitivity_vs_fdr_iterated.pdf}
  \vspace{-0.3\baselineskip}
  \caption{Simulation study 2: Complete variable selection sensitivity against false discovery rate based on 100 data simulations with one standard deviation error bars. The reference approach improves sensitivity and reduces false discovery rate for all methods.}
  \label{fig:sensitivity_vs_fdr_iterated}
\end{figure}
\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/stability_iterated.pdf}
  \vspace{-0.3\baselineskip}
  \caption{Simulation study 2: Complete variable selection stability estimates with 95\% intervals based on 100 data simulations. The reference approach improves stability for all methods.}
  \label{fig:stability_iterated}
\end{figure}
